---
title: "team.Rmd"
author: "Elisabeth Balke"
date: "10/11/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Part one: Cleaning data
Although the three Excel sheets are loaded into R, they are not useful before some cleaning. Manipulate the three imported data frames, so that in the cleaned dataset: 

1. The player names and opponent names are factors.
```{r}
library(readxl)
library(ggplot2)
library(dplyr)
defense <- read_excel('cyclonesFootball2019.xlsx', sheet='Defensive')
str(defense)

offensive <- read_excel('cyclonesFootball2019.xlsx', sheet='Offensive')
str(offensive)

biography <- read_excel('cyclonesFootball2019.xlsx', sheet='Biography')
str(biography)

#1
defense$Name <- as.factor(defense$Name)
defense$Opponent_Opponent <- as.factor(defense$Opponent_Opponent)

offensive$Name <- as.factor(offensive$Name)
offensive$Opponent_Opponent <- as.factor(offensive$Opponent_Opponent)

biography$Name <- as.factor(biography$Name)
```
2. All offensive and defensive statistics are numerical. The `Weight` column in biogrpahy is also numeric. (Optional: Learn to mutate multiple columns systematically using the `?across` function in `dplyr)
```{r}
defClean <- defense %>%
  mutate_at(vars(Tackles_Solo,Tackles_ASST,Tackles_TFL,Tackles_Sack,Turnover_FF,Turnover_FR,Turnover_INT,Pass_QBH,Pass_PB), funs(as.numeric))

offClean <- offensive %>%
  mutate_at(vars(Receiving_REC,Receiving_YDS,Receiving_TD,Rushing_ATT,Rushing_YDS,Rushing_TD,'Passing_CMP-ATT',Passing_YDS,Passing_TD,Passing_INT), funs(as.numeric))


```
3. Change the `Height` column in biogrpahy into numeric. (Hint: Use a unit that works well)
By cleaning the three data frames in R, at the end of this step you should have the following three clean data frames: `defClean`, `offClean`, and `bioClean`. Print the structures of these data frames.
```{r}
bioClean <- biography %>%
  mutate(Height = as.numeric(Height))

str(defClean)
str(offClean)
str(bioClean)
```
## Part two: Tidying
For the following questions, work on the cleaned data frames `defClean`, `offClean`, and `bioClean`.
1. Reformat the `defClean` data frame into a tidy format using `pivot_longer`. The type of statistic (Tackles_Solo, Tackles_ASST, etc) is added as a new key column named `stat`. 
```{r}
defCleanlonger <- defClean %>%
  pivot_longer(c(-Name,-Opponent_Opponent), names_to='stat',values_to='value')
```
2. Compare the distributions of the defensive statistics. What defensive skills are rare?
```{r}
#Numerical
defCleanlonger %>%
  group_by(stat) %>%
  summarise(statistics = sum(value)) %>%
  arrange(statistics)

#Graphical
ggplot(defCleanlonger, aes(stat, value)) + geom_boxplot()

#Turnover_INT seems to be the most rare defensive stat.

```
3. Did ISU have better defense against Iowa or Notre Dame? Answer this question by creating a scatterplot with x- and y-axes being the number of solo tackles (of each player). A large number of solo tackles is an indicator of good defense. 
  Now turn our attention to the player biography.
```{r}
defISUxNDxIA <- defCleanlonger %>% 
  filter(stat=='Tackles_Solo' & Opponent_Opponent %in% c('Notre Dame', 'Iowa')) %>%
  pivot_wider(names_from = Opponent_Opponent, values_from = value)
dat %>% ggplot(aes(x=Iowa, y=`Notre Dame`)) + geom_point()

#It appears defense did better in the Notre Dame game. There is a greater frequency of these tackles at higher numbers here. 

```
4. Separate the `Hometown` column into two columns, namely the city and the state. Print the first few rows of these two columns. (Hint: look at the `sep=` argument of `separate`)
```{r}
bioClean <- bioClean %>% separate(Hometown, c('City', 'State'), sep=', ')
head(bioClean)
```
5. How many players are there from each state?
```{r}
PlayerHome <- bioClean %>%
  group_by(State) %>%
  summarize(count=n()) %>%
  arrange(count)

print(PlayerHome)
```

## Part three: Joining data frames

2. Find which states do the ISU defensive and offensive players come from. Create an appropriate summary.
```{r}
#1 Question is unclear, as this question can be answered by Part Two #5. 
print(PlayerHome)

#Separated into Def and Off:
defHome <- defClean %>%
  select(Name) %>%
  group_by(Name)%>%
  distinct() %>%
  left_join(bioClean, by = "Name")

defHome <- defHome %>%
  group_by(State) %>%
  summarize(count=n()) %>%
  arrange(count)

offHome <- offClean %>%
  select(Name) %>%
  group_by(Name)%>%
  distinct() %>%
  left_join(bioClean, by = "Name")

offHome <- defHome %>%
  group_by(State) %>%
  summarize(count=n()) %>%
  arrange(count)

print(defHome)
print(offHome)
```
2. How does the performance of quarterback Brock Purdy relate to the performance of the ISU team?
```{r}
dat2 <- offClean %>%
  group_by(Opponent_Opponent)%>%
  summarise(Rushing_YDS = sum(Rushing_YDS, na.rm = TRUE), Rushing_TD = sum(Rushing_TD, na.rm = TRUE))
datPurdy <- offClean %>%
  filter(Name == 'Purdy, Brock') %>%
  select(Opponent_Opponent, Passing_YDS)

datPurdyxTeam <- dat2 %>%
  left_join(datPurdy)

print(datPurdyxTeam)

ggplot(datPurdyxTeam, aes(x = Passing_YDS, Rushing_YDS, shape = factor(Rushing_TD))) + geom_point(na.rm=TRUE)

#As passing yards increase, so do rushing yards. A positive correlation between Purdy's performance (passing) and the teams' (rushing).
```

3. Which ISU player(s) made big progress over the last year? Compare the 2019 and the 2018 data [here](../slides/cyclonesFootball2018.xlsx). 
```{r}
defense2 <- read_excel('cyclonesFootball2018.xlsx', sheet='Defensive')
str(defense2)

offensive2 <- read_excel('cyclonesFootball2018.xlsx', sheet='Offensive')
str(offensive2)

biography2 <- read_excel('cyclonesFootball2018.xlsx', sheet='Biography')
str(biography2)

defense2$Name <- as.factor(defense2$Name)
defense2$Opponent_Opponent <- as.factor(defense2$Opponent_Opponent)

offensive2$Name <- as.factor(offensive2$Name)
offensive2$Opponent_Opponent <- as.factor(offensive2$Opponent_Opponent)

biography2$Name <- as.factor(biography2$Name)

defClean2 <- defense2 %>%
  mutate_at(vars(Tackles_Solo,Tackles_ASST,Tackles_TFL,Tackles_Sack,Turnover_FF,Turnover_FR,Turnover_INT,Pass_QBH,Pass_PB), funs(as.numeric))

offClean2 <- offensive2 %>%
  mutate_at( vars(Receiving_REC,Receiving_YDS,Receiving_TD,Rushing_ATT,Rushing_YDS,Rushing_TD,'Passing_CMP-ATT',Passing_YDS,Passing_TD,Passing_INT), funs(as.numeric))

bioClean2 <- biography2 %>%
  mutate(Height = as.numeric(Height))

#Join
dat2 <- offClean %>%
  group_by(Name)%>%
  summarise(Rushing_YDS = sum(Rushing_YDS, na.rm = TRUE), Rushing_TD = sum(Rushing_TD, na.rm = TRUE))
dat2 <- dat2 %>%
  mutate(year=2019)

dat3 <- offClean2 %>%
  group_by(Name)%>%
  summarise(Rushing_YDS = sum(Rushing_YDS, na.rm = TRUE), Rushing_TD = sum(Rushing_TD, na.rm = TRUE))
dat3 <- dat3%>%
  mutate(year=2018)

Comparison <- dat2 %>%
  full_join(dat3)%>%
arrange(Comparison)

ggplot(Comparison, aes(Name, Rushing_TD, shape = factor(year))) + geom_point()

```

